---
title: "Study 4: where Olmo 3's contrastive frames enter its training"
subtitle: "englishRegisterStudy — results from the pre-registered analysis"
author: "Cesaire Tobias"
date: "2026-09-29"
format:
  pdf:
    date-format: "DD MMMM, YYYY"
    toc: true
    toc-depth: 2
    number-sections: true
    colorlinks: true
    documentclass: article
    geometry: margin=1in
    fontsize: 11pt
    linestretch: 1.5
    cite-method: citeproc
    bibliography: ../references.bib
    csl: ../harvard-cite-them-right.csl
    link-citations: true
    fig-pos: 'H'
    tbl-pos: 'H'
    include-in-header:
      text: |
        \usepackage{float}
        \usepackage{booktabs}
        \floatplacement{table}{H}
---

```{r}
#| label: setup
#| include: false
#| message: false
#| warning: false

# Every number in this document comes from the files R/analyse-study4.R
# writes, or from the committed counts in data/study4/, so it cannot drift
# from the analysis. Run `Rscript R/analyse-study4.R` from the repository root
# before rendering.
suppressPackageStartupMessages({
  library(dplyr)
  library(knitr)
})

root <- ".."
D <- file.path(root, "data", "study4")
derived <- c("confirmatory", "secondary", "midtraining", "sft-datasets",
             "dpo-models", "per-frame", "length")
paths <- file.path(D, paste0("results-", derived, ".csv"))
if (!all(file.exists(paths))) {
  stop("run R/analyse-study4.R from the repository root first; missing: ",
       paste(derived[!file.exists(paths)], collapse = ", "))
}
r <- setNames(lapply(paths, readr::read_csv, show_col_types = FALSE), derived)

read4 <- function(name) {
  readr::read_csv(file.path(D, name), show_col_types = FALSE)
}
counts <- read4("counts.csv")
generated <- read4("units-generated.csv")
stage1 <- read4("units-stage1.csv")
midtraining <- read4("units-midtraining.csv")
sft <- read4("units-sft.csv.gz")
dpo <- read4("units-dpo.csv.gz")

# The registered estimator, for each source's family rate.
source(file.path(root, "R", "ratio.R"))

FAMILY <- c("F06", "F07", "F08", "F09")
ARMS <- tibble::tribble(
  ~arm, ~checkpoint,                       ~prompt,
  "B1", "end of stage 1",                  "raw",
  "B2", "end of midtraining",              "raw",
  "B3", "final base",                      "raw",
  "N1", "Gen-QA mix from 2T",              "raw",
  "N2", "math-code-thinking mix from 2T",  "raw",
  "N3", "Round 5 mix from 2T",             "raw",
  "P1", "SFT",                             "chat template",
  "P2", "DPO",                             "chat template",
  "P3", "final instruct",                  "chat template")
MAIN_LINE <- c("B1", "B2", "B3", "P1", "P2", "P3")

fmt_ratio <- function(x) sprintf("%.2f", x)
fmt_p <- function(p) ifelse(p < 0.001, "< 0.001", sprintf("%.3f", p))
ratio_ci <- function(ratio, lower, upper) {
  sprintf("%s [%s, %s]", fmt_ratio(ratio), fmt_ratio(lower), fmt_ratio(upper))
}
ci_of <- function(row) ratio_ci(row$ratio, row$lower, row$upper)
fmt_rate <- function(x) sprintf("%.1f", x)
rate_ci <- function(rate, lower, upper) {
  sprintf("%s [%s, %s]", fmt_rate(rate), fmt_rate(lower), fmt_rate(upper))
}
rate_or_none <- function(hits, rate, lower, upper) {
  ifelse(hits == 0, "no hits", rate_ci(rate, lower, upper))
}
and_list <- function(x) {
  if (length(x) < 2) return(x)
  paste(paste(head(x, -1), collapse = ", "), "and", tail(x, 1))
}
pct <- function(x) sprintf("%.0f%%", 100 * x)
big <- function(x) format(x, big.mark = ",", scientific = FALSE, trim = TRUE)
millions <- function(x) sprintf("%.1f million", x / 1e6)
number_word <- function(n) {
  c("no", "one", "two", "three", "four", "five", "six", "seven", "eight",
    "nine", "ten")[n + 1]
}

# The generated text.
run_tbl <- generated %>%
  group_by(arm) %>%
  summarise(texts = n(), zero = sum(words == 0),
            cut = sum(finish_reason == "length"), words = sum(words),
            .groups = "drop") %>%
  inner_join(ARMS, by = "arm") %>%
  mutate(arm = factor(arm, levels = ARMS$arm)) %>%
  arrange(arm)
share_cut <- run_tbl$cut / run_tbl$texts
zero_arms <- as.character(run_tbl$arm[run_tbl$zero > 10])

# The training text. Every redacted document is in the olmOCR source.
olmocr <- midtraining %>% filter(grepl("OLMOCR", source))
redacted <- sum(olmocr$redacted)
redacted_docs <- sum(olmocr$documents)

# Each source's family rate, by the registered ratio estimator, with a
# normal interval as R/analyse-study4.R gives its rate tables.
hits_of <- function(d, frames = FAMILY, suffix = "") {
  rowSums(as.matrix(d[, paste0(frames, suffix)]))
}
rate_row <- function(key, label, kind, h, w) {
  x <- ratio_rate(h, w)
  se <- sqrt(x$var)
  tibble(key = key, source = label, kind = kind, units = x$n,
         words = x$words, hits = x$hits, rate = x$rate * 1e6,
         lower = max(0, x$rate - 1.96 * se) * 1e6,
         upper = (x$rate + 1.96 * se) * 1e6)
}
arm_row <- function(a) {
  d <- filter(generated, arm == a)
  rate_row(a, sprintf("%s, %s", a, ARMS$checkpoint[ARMS$arm == a]),
           "model output", hits_of(d), d$words)
}
ladder <- bind_rows(
  rate_row("stage1", "Stage-1 sample", "training text", hits_of(stage1),
           stage1$words),
  arm_row("B1"),
  rate_row("midtraining", "Midtraining sample", "training text",
           hits_of(midtraining), midtraining$words),
  arm_row("B2"),
  arm_row("B3"),
  rate_row("sft", "SFT responses", "training text", hits_of(sft), sft$words),
  arm_row("P1"),
  rate_row("dpo_chosen", "DPO chosen responses", "training text",
           hits_of(dpo, suffix = "_chosen"), dpo$words_chosen),
  rate_row("dpo_rejected", "DPO rejected responses", "training text",
           hits_of(dpo, suffix = "_rejected"), dpo$words_rejected),
  arm_row("P2"),
  arm_row("P3"),
  arm_row("N1"),
  arm_row("N2"),
  arm_row("N3"))
rate_of <- function(k) ladder$rate[ladder$key == k]
training_text <- ladder %>% filter(kind == "training text")

# The confirmatory tests.
conf <- r$confirmatory
cf <- function(h, o = "family") conf[conf$hypothesis == h & conf$outcome == o, ]
composite_sd <- conf$sd_frame_source[conf$outcome == "composite"][1]
composite_significant <- conf %>% filter(outcome == "composite", p_holm < 0.05)
CONTRASTS <- c(
  H1 = "B1 against the stage-1 sample",
  H2 = "Midtraining sample against the stage-1 sample",
  H3 = "B2 against B1",
  H4 = "N1 against N2",
  H5 = "SFT responses against B3",
  H6 = "P1 against B3",
  H7 = "Chosen against rejected DPO responses",
  H8 = "P2 against P1",
  H9 = "P3 against P2")

# The secondary analyses.
sec <- r$secondary
sc <- function(h, o = "family") sec[sec$hypothesis == h & sec$outcome == o, ]
sensitivity <- sec %>%
  filter(outcome == "family with F10") %>%
  inner_join(conf %>% filter(outcome == "family") %>%
               select(hypothesis, family_ratio = ratio), by = "hypothesis")
replication <- sec %>% filter(hypothesis == "replication")

# The multiple Studies 2 and 3 reported, over the frames they shared with
# Study 1, against the registered family's.
summed <- function(s, fr) {
  x <- counts %>% filter(source == s, frame_id %in% fr)
  sum(x$hits) / x$words[1]
}
STUDY23_FAMILY <- c("F07", "F08", "F09", "F10")
multiple_23 <- summed("B3", STUDY23_FAMILY) / summed("stage1", STUDY23_FAMILY)
multiple_4 <- summed("B3", FAMILY) / summed("stage1", FAMILY)

# Per frame.
pf <- r$`per-frame`
pf_rate <- function(s, f) pf$rate[pf$source == s & pf$frame_id == f]
f09_share <- function(s) {
  x <- pf %>% filter(source == s, frame_id %in% FAMILY)
  x$hits[x$frame_id == "F09"] / sum(x$hits)
}
base_arms <- c("B1", "B2", "B3", "N1", "N2", "N3")
f10_base <- sapply(base_arms, pf_rate, f = "F10")
f10_post <- sapply(c("P1", "P2", "P3"), pf_rate, f = "F10")
along <- function(f) sapply(MAIN_LINE, pf_rate, f = f)
near_stage1 <- function(f) all(abs(log(along(f) / pf_rate("stage1", f))) < log(1.5))
FRAME_SOURCES <- c("stage1", MAIN_LINE)

# The training text, broken down.
mt <- r$midtraining
mt_cat <- mt %>% filter(level == "category") %>% arrange(desc(per_million))
mt_src <- mt %>% filter(level == "source") %>% arrange(desc(per_million))
cat_rate <- function(c) mt_cat$per_million[mt_cat$category == c]
src_rate <- function(s) mt_src$per_million[mt_src$source == s]
sft_sets <- r$`sft-datasets` %>% arrange(desc(per_million))
wildchat <- sft_sets %>% filter(source_dataset == "Wildchat")
dm <- r$`dpo-models`
largest <- function(s) dm %>% filter(side == s) %>% arrange(desc(units)) %>% slice(1)

# Rates by how each text ended.
len <- r$length
len_rate <- function(a, reason) {
  len$per_million[len$arm == a & len$finish_reason == reason]
}
ordered_within <- function(reason) {
  all(diff(sapply(MAIN_LINE, len_rate, reason = reason)) > 0)
}
finished_higher <- sum(sapply(ARMS$arm, function(a) {
  len_rate(a, "stop") > len_rate(a, "length")
}))

# The prose below states these in words; stop rather than let it drift.
stopifnot(
  all(run_tbl$texts == 4000),
  cf("H1")$verdict == "evidence against",
  cf("H2")$verdict == "evidence against",
  cf("H3")$verdict == "supported",
  grepl("N1", cf("H4")$verdict),
  cf("H5")$verdict == "evidence against",
  all(sapply(c("H6", "H7", "H8", "H9"), function(h) {
    cf(h)$verdict == "supported"
  })),
  nrow(composite_significant) == 1,
  composite_significant$hypothesis == "H5",
  composite_significant$ratio < 1,
  sc("long-context stage")$lower > 1,
  rate_of("B1") < rate_of("stage1"),
  max(training_text$upper) <
    min(ladder$lower[ladder$key %in% c("B2", "B3", "P1", "P2", "P3")]),
  sc("midtrained model vs its data")$lower > 1,
  sc("SFT model vs its data")$lower > 1,
  sc("midtrained model vs its data", "composite")$lower > 1,
  sc("SFT model vs its data", "composite")$lower > 1,
  sc("Round 5 against Gen-QA")$upper < 1,
  sc("Round 5 against math-code")$lower > 1,
  all(replication$lower < 1 & replication$upper > 1),
  all(sensitivity$lower > 1 | sensitivity$upper < 1),
  all(sign(log(sensitivity$ratio)) == sign(log(sensitivity$family_ratio))),
  sensitivity$ratio[sensitivity$hypothesis == "H3"] < cf("H3")$ratio,
  all(abs(log(f10_base / pf_rate("stage1", "F10"))) < log(1.5)),
  min(f10_post) > max(f10_base),
  all(diff(along("F07")[c("B1", "B2", "B3")]) > 0),
  names(which.max(along("F08"))) == "B3",
  all(c(along("F01")[c("P1", "P3")], along("F03")[c("P1", "P3")]) >
        3 * c(along("F01")["B3"], along("F01")["B3"],
              along("F03")["B3"], along("F03")["B3"])),
  all(sapply(c("F01", "F03"), function(f) {
    x <- along(f)
    x[["P1"]] / x[["B3"]] > max(x[["P2"]] / x[["P1"]], x[["P3"]] / x[["P2"]])
  })),
  along("F04")["P3"] < pf_rate("stage1", "F04") / 2,
  near_stage1("F13"), near_stage1("F14"),
  max(pf$rate[pf$frame_id == "F06"]) < 5,
  ordered_within("stop"), ordered_within("length"),
  mt_cat$category[1] == "Web pages",
  mt_cat$hits[1] / sum(mt_cat$hits) > 0.5,
  all(sapply(c("QA (synth)", "Instruction (synth)"), cat_rate) <
        cat_rate("Web pages")),
  all(sapply(c("QA (synth)", "Instruction (synth)"), cat_rate) >
        max(sapply(c("Math (synth)", "Code", "Python (synth)",
                     "Thinking (synth)", "PDFs"), cat_rate))),
  mt_src$source[1] == "Nemotron Synth QA",
  sft_sets$source_dataset[1] == "Wildchat",
  wildchat$hits / sum(sft_sets$hits) > 0.5,
  sum(midtraining$redacted) == redacted,
  all(zero_arms %in% c("B1", "B2", "N1", "N3")),
  sum(run_tbl$zero[run_tbl$arm %in% c("P1", "P2", "P3")]) == 0
)
```

# What this measures

Studies 2 and 3 of this project found, in analyses registered as exploratory,
that Olmo 3 7B writes the contrastive frames, *it's not X, it's Y* and its
relatives, at several times the rate of the web text it was pretrained on, and
that the instruction-tuned model writes them more often again. Study 4 asks
where in the model's training that excess enters. It was registered at
<https://osf.io/d79u4> on 29 September 2026, before any of its text was
generated or any frame counted in its data.

Olmo 3 was trained in six stages: stage-1 pretraining, midtraining,
long-context training, supervised fine-tuning (SFT), preference tuning (DPO)
and reinforcement learning with verifiable rewards (RLVR). Ai2 published a
checkpoint after each stage and the data each stage trained on
[@olmo2025olmo]. Study 4 counts the frames in the output of checkpoints from
each stage and in the text four of the stages trained on.

```{r}
#| label: tbl-arms
#| echo: false
ARMS %>%
  transmute(Arm = arm, Checkpoint = checkpoint, Prompt = prompt) %>%
  kable(caption = paste("The nine checkpoints. B1 to N3 are branches of",
                        "allenai/Olmo-3-1025-7B, and P1 to P3 are",
                        "allenai/Olmo-3-7B-Instruct-SFT,",
                        "allenai/Olmo-3-7B-Instruct-DPO and",
                        "allenai/Olmo-3-7B-Instruct. Each is pinned to a",
                        "commit in docs/study4-protocol.md."))
```

Each checkpoint wrote one post for each of the 4,000 titles Study 2 used,
drawn from GloWbE's blog pages [@davies2013glowbe] and prompted with
`Write a blog post titled:` and the title, at temperature 1.0 and top-p 1.0
with at most 1,024 new tokens. The base-model checkpoints received the prompt
as raw text and the three post-trained ones through their chat template, which
adds a default system prompt, as in Study 2. N1 to N3 are three midtraining
runs of the same length from one stage-1 checkpoint at 2 trillion tokens, which
Ai2 published as branches of the base model, each with a different data mix.
In the Olmo 3 report's description, the Gen-QA mix "increases proportions of
web, QA, and instruction data while omitting math, code, and thinking", the
math-code-thinking mix "increases proportions of math, code, and thinking data
while omitting QA and instruction data" and keeps web, and Round 5 is the final
mix [@olmo2025olmo].

Four samples of training text are measured:

- **Stage 1:** Study 3's registered sample of `common_crawl`, the web portion
  of the stage-1 mix, `r big(nrow(stage1))` files and
  `r millions(sum(stage1$words))` words.
- **Midtraining:** `r millions(sum(midtraining$words))` words from
  `r big(nrow(midtraining))` files of the Dolmino midtraining mix, allocated
  across its 24 sources by their shares of its tokens, so the pooled sample
  stands for the mix.
- **SFT:** the assistant turns of all `r big(nrow(sft))` conversations in the
  SFT set, `r millions(sum(sft$words))` words.
- **DPO:** the final response on each side of all `r big(nrow(dpo))`
  preference pairs, `r millions(sum(dpo$words_chosen))` words chosen and
  `r millions(sum(dpo$words_rejected))` rejected.

The long-context stage's data, two-thirds of it midtraining data, is not
counted. The RLVR data holds prompts and reference answers but none of the
model's responses, so RLVR is measured on the output side only.

## Two outcomes

The **family** is the four contrastive frames F06 to F09, `it 's not * , it 's`,
`it 's not just`, `not just * but` and `is n't just`, as a rate per word.
Each source's rate is its total hits over its total words, with an interval
from the variation between its units, whether posts, files, conversations or
pairs (`R/ratio.R`). The **composite** is all fifteen frozen frames through the
model Studies 2 and 3 used, in which a source's effect is its frame rates
averaged over frames on the log scale, so that it means what it meant in
Studies 1 to 3.

Studies 2 and 3 summarised the contrastive family over the frames they shared
with Study 1, F07 to F10. F10, `not * , but *`, is a partial frame, and Study 4
registered the family without it, adding it only as a sensitivity check. The
two sets give different multiples for the same text. The final base model's
rate is `r fmt_ratio(multiple_23)` times the stage-1 sample's over F07 to F10,
the figure behind the registration's "about five times", and
`r fmt_ratio(multiple_4)` times over F06 to F09. F10 runs near the stage-1
sample's rate in every base-model checkpoint, so including it pulls the
multiple towards 1.

## Nine hypotheses

- **H1.** The end-of-stage-1 checkpoint (B1) uses the frames more than its
  stage-1 input.
- **H2.** The midtraining text uses them more than the stage-1 text.
- **H3.** The end-of-midtraining checkpoint (B2) uses them more than B1.
- **H4.** The Gen-QA checkpoint (N1) and the math-code-thinking checkpoint
  (N2) differ, in either direction.
- **H5.** The SFT responses use them more than the final base model (B3).
- **H6.** The SFT checkpoint (P1) uses them more than B3.
- **H7.** Chosen DPO responses use them more than rejected responses to the
  same prompts.
- **H8.** The DPO checkpoint (P2) uses them more than P1.
- **H9.** The final instruct model (P3) uses them more than P2.

Each is tested on both outcomes, eighteen tests with Holm's correction. A
hypothesis is supported on an outcome when its ratio lies in the predicted
direction with an adjusted p below 0.05, and a ratio in the other direction
with an adjusted p below 0.05 is evidence against it.

Earlier work measured post-training's effect on style from output alone.
@reinhart2025llms compared Llama 3 and GPT-4o with human writing on Biber's
grammatical and rhetorical features and found the differences larger for
instruction-tuned models than for base models, and @li2026narrative compared
OLMo 32B's base, SFT, DPO and RLVR checkpoints with human fiction and found
that post-training compresses stylistic variation. Study 4 adds the stages
before post-training and the training text of each stage.

## Deviations

Four are recorded in `docs/deviations.md`, and none changes the design:

- **S4-D1:** the protocol's directory pattern for one midtraining source
  matched nothing and is read as the one pattern that fits.
- **S4-D2:** the registration gave the DPO set as 260,000 pairs, the dataset
  card's rounded description. It holds `r big(nrow(dpo))`, and every pair was
  counted.
- **S4-D3:** SFT conversations are identified by file and row, where the
  protocol names each conversation's `id`.
- **S4-D4:** `r big(redacted)` of the `r big(redacted_docs)` documents drawn
  from the midtraining mix's olmOCR source were redacted after training and are
  counted as they stand, as in Study 3.

S4-D1, S4-D2 and S4-D4 were filed as an update to the registration, approved
before the analysis ran.

# The generated text

```{r}
#| label: tbl-run
#| echo: false
run_tbl %>%
  transmute(Arm = as.character(arm),
            Words = big(words),
            `Words a post` = sprintf("%.0f", words / texts),
            `Reached 1,024 tokens` = sprintf("%s (%s)", big(cut),
                                             pct(cut / texts)),
            `No words` = zero) %>%
  kable(align = c("l", "r", "r", "r", "r"),
        caption = "The nine arms as generated on 29 September 2026.")
```

Every arm wrote all 4,000 posts. Between `r pct(min(share_cut))` and
`r pct(max(share_cut))` of an arm's posts reached the 1,024-token limit and
end mid-text; they are counted as they came, as registered, and the length
analysis below shows what that does to the rates. Posts with no words are ones
the model ended at once or that hold no word character, and they count as zero
words. They cluster in `r and_list(zero_arms)`, and the post-trained
checkpoints wrote none. The checks made on the text and the counts
before the analysis are in `docs/collection-log.md`.

# Confirmatory results

## The family

```{r}
#| label: tbl-family
#| echo: false
conf %>%
  filter(outcome == "family") %>%
  transmute(` ` = hypothesis,
            Contrast = CONTRASTS[hypothesis],
            Ratio = ratio_ci(ratio, lower, upper),
            `Holm p` = fmt_p(p_holm),
            Verdict = ifelse(hypothesis == "H4", "favours N1", verdict)) %>%
  kable(caption = paste("The family outcome: the first source's rate of F06",
                        "to F09 over the second's, with unadjusted 95%",
                        "intervals. H4 is two-sided; a ratio above 1 favours",
                        "the QA and instruction data in N1's mix."))
```

The end-of-stage-1 checkpoint uses the family less than its training text: B1's
rate is `r ci_of(cf("H1"))` times the stage-1 sample's, evidence against H1.
The midtraining text also carries less of it than the stage-1 text,
`r ci_of(cf("H2"))`, evidence against H2. Midtraining still raises the model's
rate: B2 is `r ci_of(cf("H3"))` times B1, which supports H3. Of the two
midtraining mixes run from the same checkpoint, the one with more web, QA and
instruction data and no math, code or reasoning traces gives the higher rate:
N1 is `r ci_of(cf("H4"))` times N2, which favours the QA and instruction data.
The mixes differ in several categories at once, so the attribution to QA and
instruction data is an interpretation.

The SFT responses use the family at `r ci_of(cf("H5"))` times the final base
model's rate, evidence against H5, yet the SFT checkpoint uses it
`r ci_of(cf("H6"))` times as often as B3. The preference data leans the same
way as the model. Chosen responses use the family `r ci_of(cf("H7"))` times as often as
the rejected responses to the same prompts. DPO then raises the model's rate by
`r ci_of(cf("H8"))` and RLVR by a further `r ci_of(cf("H9"))`. H6 to H9 are
supported.

## The composite

```{r}
#| label: tbl-composite
#| echo: false
conf %>%
  filter(outcome == "composite") %>%
  transmute(` ` = hypothesis,
            Contrast = CONTRASTS[hypothesis],
            Ratio = ratio_ci(ratio, lower, upper),
            `Holm p` = fmt_p(p_holm),
            Verdict = verdict) %>%
  kable(caption = paste("The composite outcome: the first source's effect over",
                        "the second's in the model over all fifteen frames,",
                        "with unadjusted 95% intervals."))
```

On the composite only H5 reaches significance, again as evidence against: the
SFT responses' effect is `r ci_of(cf("H5", "composite"))` times B3's. Every
other contrast is not supported. The frame-by-source SD is
`r sprintf("%.2f", composite_sd)` on the log scale. Along the training line
some frames rise steeply, others hold steady and F04 falls, so an average that
weights each of the fifteen equally on the log scale moves little and carries
wide intervals. The per-frame rates below show which frames move.

## The family along the training line

```{r}
#| label: tbl-ladder
#| echo: false
ladder %>%
  transmute(Source = source,
            Kind = kind,
            Words = big(words),
            Hits = big(hits),
            `Per million` = rate_ci(rate, lower, upper)) %>%
  kable(align = c("l", "l", "r", "r", "r"),
        caption = paste("The family's rate per million words in each source,",
                        "in training order, with the midtraining experiment",
                        "last. Intervals are rate plus or minus 1.96",
                        "standard errors from the ratio estimator."))
```

The model's rate rises at every stage after stage 1: by a factor of
`r fmt_ratio(cf("H3")$ratio)` in midtraining,
`r fmt_ratio(sc("long-context stage")$ratio)` in long-context training (a
secondary contrast), `r fmt_ratio(cf("H6")$ratio)` in SFT,
`r fmt_ratio(cf("H8")$ratio)` in DPO and `r fmt_ratio(cf("H9")$ratio)` in RLVR.
From B1's `r fmt_rate(rate_of("B1"))` per million words it reaches
`r fmt_rate(rate_of("P3"))` in the final instruct model. None of the training
text comes near the model's rate once midtraining has run: it runs between
`r fmt_rate(min(training_text$rate))` and `r fmt_rate(max(training_text$rate))`
per million, against `r fmt_rate(rate_of("B2"))` for B2.

# Secondary results

Everything in this section is registered as secondary and reported with
unadjusted intervals.

```{r}
#| label: tbl-secondary
#| echo: false
sec %>%
  filter(hypothesis %in% c("long-context stage", "Round 5 against Gen-QA",
                           "Round 5 against math-code",
                           "midtrained model vs its data",
                           "SFT model vs its data")) %>%
  select(hypothesis, plus, minus, outcome, ratio, lower, upper) %>%
  mutate(value = ratio_ci(ratio, lower, upper)) %>%
  select(hypothesis, plus, minus, outcome, value) %>%
  tidyr::pivot_wider(names_from = outcome, values_from = value) %>%
  transmute(Contrast = sprintf("%s against %s", plus,
                               ifelse(minus == "midtraining",
                                      "the midtraining sample",
                                      ifelse(minus == "sft", "the SFT responses",
                                             minus))),
            Family = family, Composite = composite) %>%
  kable(caption = "Secondary contrasts, with unadjusted 95% intervals.")
```

Long-context training raises the family's rate by
`r ci_of(sc("long-context stage"))`, and by
`r ci_of(sc("long-context stage", "composite"))` on the composite. Its data,
two-thirds of it midtraining data, was not counted, so this study cannot say
what in that stage raises the rate.

The Round 5 mix from 2T lands between the other two: N3 is
`r ci_of(sc("Round 5 against Gen-QA"))` times N1 and
`r ci_of(sc("Round 5 against math-code"))` times N2.

The two checkpoints set against the text of their own stage both outrun it. B2
is `r ci_of(sc("midtrained model vs its data"))` times the midtraining sample
and P1 is `r ci_of(sc("SFT model vs its data"))` times the SFT responses, and
both hold on the composite.

Adding the partial F10 to the family leaves every confirmatory ratio on the
same side of 1, with an interval that excludes 1. Where F10 makes up much of
the hits and changes little between the two sources, it pulls the ratio
towards 1: H3's falls from `r fmt_ratio(cf("H3")$ratio)` to
`r fmt_ratio(sensitivity$ratio[sensitivity$hypothesis == "H3"])`.

The final base and final instruct checkpoints are the ones Study 2 ran at the
same settings, with new seeds, and their family rates reproduce Study 2's. B3
is `r ci_of(replication[replication$plus == "B3", ])` times Study 2's
base-1.0, and P3 is `r ci_of(replication[replication$plus == "P3", ])` times
its instruct-1.0.

## Where the family sits in the training text

```{r}
#| label: tbl-midtraining
#| echo: false
mt_cat %>%
  transmute(Category = category,
            Files = units,
            Words = millions(words),
            Hits = big(hits),
            `Per million` = rate_or_none(hits, per_million, lower, upper)) %>%
  kable(align = c("l", "r", "r", "r", "r"),
        caption = paste("The midtraining sample by the dataset card's",
                        "categories, with normal intervals."))
```

In the midtraining sample, `r pct(mt_cat$hits[1] / sum(mt_cat$hits))` of the
family's hits come from web text, which runs at
`r fmt_rate(cat_rate("Web pages"))` per million, close to the stage-1 sample's
rate. The synthetic QA sources run at `r fmt_rate(cat_rate("QA (synth)"))` as a
group and the instruction data at `r fmt_rate(cat_rate("Instruction (synth)"))`,
with one QA source, Nemotron Synth QA, the highest of the 24 at
`r fmt_rate(src_rate("Nemotron Synth QA"))`. Code, mathematics, reasoning
traces and PDFs carry little. A mix that trades mathematics, code and reasoning
traces for QA and instruction data, as the Gen-QA mix does, therefore holds
more of the family per word, which may account for part of N1's higher rate.
The two experimental mixes' own text was not measured, and the released mix,
which was, differs from both.

```{r}
#| label: tbl-sft
#| echo: false
sft_sets %>%
  transmute(Dataset = source_dataset,
            Conversations = big(units),
            Words = big(words),
            Hits = big(hits),
            `Per million` = rate_or_none(hits, per_million, lower, upper)) %>%
  kable(align = c("l", "r", "r", "r", "r"),
        caption = paste("The SFT responses by source dataset, with normal",
                        "intervals."))
```

In the SFT set the family comes mostly from one dataset: the WildChat prompts
with responses written by GPT-4.1 hold `r pct(wildchat$words / sum(sft_sets$words))`
of the words and `r pct(wildchat$hits / sum(sft_sets$hits))` of the family's
hits, at `r fmt_rate(wildchat$per_million)` per million. That is
`r pct(wildchat$per_million / rate_of("P1"))` of the SFT checkpoint's rate. The
mathematics, code and puzzle datasets carry little or none.

In the DPO set, `r largest("chosen")$model` wrote
`r pct(largest("chosen")$units / nrow(dpo))` of the chosen responses, at
`r fmt_rate(largest("chosen")$per_million)` per million, and
`r largest("rejected")$model` wrote
`r pct(largest("rejected")$units / nrow(dpo))` of the rejected ones, at
`r fmt_rate(largest("rejected")$per_million)`. H7's difference therefore
reflects which models wrote each side as well as which responses were
preferred. The per-model rates are in `data/study4/results-dpo-models.csv`.

# Exploratory results

Everything in this section is registered as exploratory. It is descriptive and
uncorrected.

## Frame by frame

```{r}
#| label: tbl-frames
#| echo: false
pf %>%
  filter(source %in% FRAME_SOURCES) %>%
  mutate(source = factor(source, levels = FRAME_SOURCES)) %>%
  select(source, frame_id, rate) %>%
  arrange(source) %>%
  tidyr::pivot_wider(names_from = source, values_from = rate) %>%
  arrange(frame_id) %>%
  mutate(across(-frame_id, fmt_rate)) %>%
  rename(Frame = frame_id, `Stage 1` = stage1) %>%
  kable(align = c("l", rep("r", length(FRAME_SOURCES))),
        caption = paste("Each frame's rate per million words in the stage-1",
                        "sample and along the main line of checkpoints. The",
                        "frames' strings are in data/frames.csv."))
```

F09, `is n't just`, carries most of the family's rise. It runs at
`r fmt_rate(pf_rate("stage1", "F09"))` per million in the stage-1 sample, at
`r fmt_rate(pf_rate("B3", "F09"))` in the final base model and at
`r fmt_rate(pf_rate("P3", "F09"))` in the final instruct model, where it makes
up `r pct(f09_share("P3"))` of the family's hits. F07, `it 's not just`, rises
at each step from B1 to B3, and F08, `not just * but`, peaks in B3. F06, the
full `it 's not * , it 's`, is rare everywhere, and F10 holds near the stage-1
rate in every base-model checkpoint and runs higher in the post-trained ones.

The other families move differently. Two announce-the-point frames rise mainly
at SFT: F01, `here 's the thing`, from `r fmt_rate(pf_rate("B3", "F01"))` per
million in B3 to `r fmt_rate(pf_rate("P3", "F01"))` in P3, and F03,
`here 's what`, from `r fmt_rate(pf_rate("B3", "F03"))` to
`r fmt_rate(pf_rate("P3", "F03"))`. F04, `the thing is`, falls from
`r fmt_rate(pf_rate("stage1", "F04"))` in the stage-1 sample to
`r fmt_rate(pf_rate("P3", "F04"))` in P3, while F13, `and that 's`, and F14,
`turns out`, stay near their stage-1 rates throughout. That spread is what the
composite's frame-by-source SD records.

## Length and truncation

```{r}
#| label: tbl-length
#| echo: false
len %>%
  select(arm, finish_reason, texts, per_million) %>%
  tidyr::pivot_wider(names_from = finish_reason,
                     values_from = c(texts, per_million)) %>%
  mutate(arm = factor(arm, levels = ARMS$arm)) %>%
  arrange(arm) %>%
  transmute(Arm = as.character(arm),
            `Finished posts` = big(texts_stop),
            `Per million, finished` = fmt_rate(per_million_stop),
            `Cut-off posts` = big(texts_length),
            `Per million, cut off` = fmt_rate(per_million_length)) %>%
  kable(align = c("l", "r", "r", "r", "r"),
        caption = paste("The family's rate per million words by whether the",
                        "post finished or reached the token limit."))
```

In `r number_word(finished_higher)` of the nine arms, posts that finished carry
the family at a higher rate than posts cut at the token limit, so truncation
lowers those arms' rates, more so where more posts were cut. The order along
the training line holds within each kind of post: among finished posts and
among cut-off posts alike, each of B1, B2, B3, P1, P2 and P3 has a higher rate
than the one before.

# What this does not establish

- **One model.** Olmo 3 7B was chosen because every stage's checkpoint and data
  are published. Nothing here says how other models acquire the frames.
- **One prompt and one genre.** Every text answers `Write a blog post titled:`
  followed by a title from a 2012 blog.
- **Genre in the comparisons with training text.** H1, H2, H5 and the
  secondary comparisons of a checkpoint with its own data set prompted blog
  posts against training text of every kind, from code to encyclopedia
  articles. The comparisons between checkpoints hold the prompts fixed. Within
  the training text, the sources nearest to blog writing, web text and GPT-4.1's
  WildChat responses, still run far below the post-trained checkpoints.
- **What a stage bundles.** A difference between successive checkpoints is the
  effect of all the training between them: the data, the learning-rate
  schedule and, at the step from base to SFT, the prompt format and the chat
  template's default system prompt.
- **One experiment on the mixes.** N1 to N3 start from a stage-1 checkpoint at
  2 trillion tokens, earlier than the one the released model continued from,
  and their mixes differ in several categories at once.
- **Stages measured on one side.** The long-context stage's data was not
  counted, and the RLVR data holds no responses. For both, only the change in
  the model's output is measured.
- **Few hits at the start.** B1's rate rests on
  `r big(ladder$hits[ladder$key == "B1"])` hits, so its interval is wide.
- **A midtraining sample by token share.** The sources are allocated by their
  shares of tokens and rates are per word, so the pooled sample matches the
  mix's composition only approximately, and the olmOCR source's rate describes
  its unredacted documents only (S4-D4).
- **Truncated texts.** Between `r pct(min(share_cut))` and
  `r pct(max(share_cut))` of an arm's posts were cut at 1,024 tokens. The
  length analysis above shows they lower the rates without changing the order.
- **Fixed strings chosen as typical of model output.** As in Studies 1 to 3,
  these are rates per million of fifteen frozen strings, chosen because they
  seem typical of machine-generated prose. Density, sentence position and
  paraphrases are not measured.
- **Text that cannot be regenerated exactly.** Each generation's seed fixes its
  sampling request, but vLLM's arithmetic varies with the batch a request
  shares, so a rerun gives different text.
  `data/study4/generated-sha256.txt` identifies the files that were counted.

# Reproducing this

The repository holds the generation and counting code
(`python/generate_study4.py`, `python/count_study4.py`), the protocol with
every checkpoint's and dataset's pinned commit (`docs/study4-protocol.md`), the
counts (`data/study4/counts.csv`, with one row per unit of text in
`data/study4/units-*`), one record per midtraining file read
(`data/study4/midtraining/`), and the record of the run
(`docs/collection-log.md`).

`Rscript R/analyse-study4.R` recomputes every number in this document from
those files, and `python python/count_study4.py --collect data/study4`
rebuilds the tables from the generated text and the per-shard counts, which
git does not hold.

# References

::: {#refs}
:::
