---
title: "Study 1: rhetorical frames across national varieties of web English"
subtitle: "englishRegisterStudy — results from the pre-registered analysis"
author: "Cesaire Tobias"
date: "2026-09-25"
format:
  pdf:
    date-format: "DD MMMM, YYYY"
    toc: true
    toc-depth: 2
    number-sections: true
    colorlinks: true
    documentclass: article
    geometry: margin=1in
    fontsize: 11pt
    linestretch: 1.5
    cite-method: citeproc
    bibliography: ../references.bib
    csl: ../harvard-cite-them-right.csl
    link-citations: true
    fig-pos: 'H'
    tbl-pos: 'H'
    include-in-header:
      text: |
        \usepackage{float}
        \usepackage{booktabs}
        \floatplacement{table}{H}
---

```{r}
#| label: setup
#| include: false
#| message: false
#| warning: false

# Every number in this document comes from the files R/analyse.R writes, so it
# cannot drift from the analysis. Run `Rscript R/analyse.R` from the repository
# root before rendering.
suppressPackageStartupMessages({
  library(dplyr)
  library(knitr)
})

root <- ".."
need <- c("composite-ratios", "composite-ratios-sensitivity",
          "composite-ratios-by-section", "composite-combined",
          "ratios-vs-us")
paths <- file.path(root, "data", paste0(need, ".csv"))
if (!all(file.exists(paths))) {
  stop("run Rscript R/analyse.R from the repository root first; missing: ",
       paste(need[!file.exists(paths)], collapse = ", "))
}
d <- setNames(lapply(paths, readr::read_csv, show_col_types = FALSE), need)

frames <- readr::read_csv(file.path(root, "data", "frames.csv"),
                          show_col_types = FALSE)

fmt_ratio <- function(x) sprintf("%.2f", x)
fmt_p <- function(p) ifelse(p < 0.0001, "< 0.0001", sprintf("%.4f", p))
ratio_ci <- function(ratio, lower, upper) {
  sprintf("%s [%s, %s]", fmt_ratio(ratio), fmt_ratio(lower), fmt_ratio(upper))
}
n_countable <- sum(frames$interface != "uncountable")
n_split <- sum(frames$interface == "split")
```

# What this measures

Fifteen rhetorical frames were frozen before any querying, chosen from an
impression of what language models overproduce: announcing a point before
making it, the *it's not X, it's Y* contrast, the epigrammatic assertion. The
question is whether they are features of American web register rather than of
English generally.

The instrument is GloWbE [@davies2013glowbe], 1.9 billion words of web text
from twenty countries, collected in December 2012. Five national components are
used: US, GB, IE, AU and ZA. The collection date is a design choice rather than
an accident. Any corpus collected after 2022 contains generated text in unknown
proportion, so a human baseline drawn from one would be circular.

The design, the frames and the analysis were registered at
<https://osf.io/48wjn> on 21 September 2026, before any count was retrieved.
Two deviations were declared during collection and filed as registration
updates; both are in `docs/deviations.md`.

# What the interface allowed

```{r}
#| label: tbl-interface
#| echo: false
frames %>%
  count(interface, name = "frames") %>%
  mutate(interface = recode(interface,
    split = "Counted in each section",
    combined_only = "Counted only with sections combined",
    uncountable = "Not countable on a free account")) %>%
  kable(col.names = c("What the interface allowed", "Frames"),
        caption = "The free tier's limits decided the shape of the analysis.")
```

A free english-corpora.org account refuses a section-restricted search when
every slot in the query occurs 1,500,000 times or more, and caps a search
string at five tokens. The first blocks six frames, which is the whole
contrastive family apart from one, plus two others. The second blocks F06,
`it 's not * , it 's`, entirely: it is seven tokens.

The consequence is deviation D2. The confirmatory analysis covers the
`r n_countable` countable frames with both sections combined, and the
registered per-section model becomes a secondary analysis on the `r n_split`
frames that split. The genre control is what this costs: it now covers
`r n_split` frames rather than fifteen, and not the frames whose rates a genre
explanation would most plausibly account for.

# Results

## The composite

```{r}
#| label: tbl-primary
#| echo: false
d$`composite-ratios` %>%
  transmute(Variety = variety,
            `Ratio against US` = ratio_ci(ratio, ratio_lower, ratio_upper),
            `Holm p` = fmt_p(p_holm),
            `Supports H1` = ifelse(supports_hypothesis, "yes", "no")) %>%
  kable(caption = paste("Confirmatory result: the composite rate of the",
                        n_countable, "countable frames in each variety,",
                        "relative to American, with 95% intervals."))
```

Every variety uses the frames less often than American English does, and every
comparison survives Holm's correction. The hypothesis was directional and
pre-registered, and all four varieties support it.

The frame-by-variety random effect has an SD of
`r sprintf("%.3f", d$"composite-ratios"$sd_frame_variety[1])` on the log scale,
so frames disagree modestly about how large the gap is. That disagreement is
carried into the intervals above, which is what the model is for.

```{r}
#| label: tbl-rates
#| echo: false
d$`composite-combined` %>%
  transmute(Variety = variety,
            Hits = format(hits, big.mark = ","),
            `Per million` = sprintf("%.1f", rate),
            `95% interval` = sprintf("[%.1f, %.1f]", lower, upper)) %>%
  kable(caption = "The same thing descriptively: all countable frames summed.")
```

## Robustness

Two frames are marked partial in `data/frames.csv` because they match text
outside the construction they stand for: `not * , but *` and `and that 's`.
The second is the larger worry, being roughly ten times the size of any other
frame in raw counts. Removing both and refitting moves every ratio slightly
further from 1 rather than towards it, so the result does not rest on what
those two patterns are really counting.

```{r}
#| label: tbl-sensitivity
#| echo: false
inner_join(
  d$`composite-ratios` %>%
    transmute(variety, primary = ratio_ci(ratio, ratio_lower, ratio_upper)),
  d$`composite-ratios-sensitivity` %>%
    transmute(variety, without = ratio_ci(ratio, ratio_lower, ratio_upper)),
  by = "variety") %>%
  kable(col.names = c("Variety", "All countable frames",
                      "Without the partial frames"),
        caption = "The primary result with and without F10 and F13.")
```

## The genre control

```{r}
#| label: tbl-secondary
#| echo: false
d$`composite-ratios-by-section` %>%
  filter(model == "common") %>%
  transmute(Variety = variety,
            `Ratio against US` = ratio_ci(ratio, ratio_lower, ratio_upper),
            `Holm p` = fmt_p(p_holm)) %>%
  kable(caption = paste("Secondary analysis: the registered per-section model",
                        "on the", n_split, "frames the interface splits."))
```

The variety × section interaction test finds no evidence that the gap differs
between the corpus's blog and general sections. That is the weak form of
evidence against a pure genre explanation: if the difference between varieties
were really a difference between blogs and broadsheets, holding the section
constant should have shrunk it, and it does not.

The control is weaker than it looks, in two ways. The corpus's own composition
table heads its General column "General (may also include blogs)", so the
sections are not clean genres; @murphy2025please abandoned balancing a GloWbE
sample across them on the strength of @biber2015exploring, a paper this study
could not obtain and therefore cites at second hand. And the control covers
`r n_split` frames, not all of them.

## Where the pattern is not uniform

```{r}
#| label: tbl-exceptions
#| echo: false
d$`ratios-vs-us` %>%
  filter(ratio > 1) %>%
  inner_join(select(frames, frame_id, query), by = "frame_id") %>%
  transmute(Frame = frame_id, Query = query, Variety = variety,
            Ratio = ratio_ci(ratio, ratio_lower, ratio_upper)) %>%
  arrange(Frame, Variety) %>%
  kable(caption = "Every per-frame cell where a variety exceeds American.")
```

```{r}
#| label: family-summary
#| include: false
fam <- d$`ratios-vs-us` %>%
  group_by(family) %>%
  summarise(median_ratio = median(ratio), n = n(), .groups = "drop") %>%
  arrange(median_ratio)
below <- sum(d$`ratios-vs-us`$ratio < 1)
total <- nrow(d$`ratios-vs-us`)
```

`r below` of the `r total` per-frame ratios are below 1. The exceptions are
British and Australian, and mostly contrastive frames. By family, the median
ratio runs from `r sprintf("%.2f", fam$median_ratio[1])` for the
`r fam$family[1]` frames to `r sprintf("%.2f", fam$median_ratio[nrow(fam)])`
for the `r fam$family[nrow(fam)]` frames: the announce-the-point moves separate
the varieties most, and the contrastive ones least. These ratios are
descriptive, uncorrected and registered as such, so this is a lead for the
write-up rather than a finding.

# What this does not establish

- **It is 2012 web English.** The register in question is claimed to have
  spread through Medium, Substack and LinkedIn, which mostly postdate the
  corpus. Nothing here speaks to English since.
- **It counts fixed strings.** The sharper version of the claim is about
  density and near-obligatory use, and a rate per million is a weak proxy for
  it. Sentence position, one-sentence paragraphs and closing epigrams are not
  reachable through the web interface at all.
- **The frames were chosen from intuition.** Freezing and registering them
  controls for choosing after seeing results. It cannot show they were the
  right fifteen.
- **Country assignment is a proxy.** GloWbE assigns pages by site rather than
  by author nationality. From a spelling test, @murphy2025please estimates
  10–15% of writers in the GB and US components are non-nationals, which biases
  a variety comparison toward the null: the gaps above are more likely
  understated than manufactured.
- **No generated text appears anywhere in this study.** That American web
  English used these frames more in 2012 says nothing on its own about why
  generated prose reads as it does. Study 2 measures model output and Study 3
  measures the training data behind it; only with those does the argument about
  machine writing become testable.

# Reproducing this

Every count came from the GloWbE web interface with a free account. The
repository holds the frozen frames (`data/frames.csv`), the counts
(`data/counts.csv` and `data/counts-combined.csv`), the section and component
word counts (`data/corpus-sizes.csv`), and what each query session did,
including the refusals (`docs/collection-log.md`).

`Rscript R/analyse.R` recomputes every number in this document.
`Rscript R/check-corpus-sizes.R` recomputes the word counts from the corpus's
own metadata download. `Rscript R/calibrate-composite.R` reruns the simulations
behind the choice of model and interval.

# References
