---
title: "Studies 2 and 3: rhetorical frames in Olmo 3's prose and its training data"
subtitle: "englishRegisterStudy — results from the pre-registered analyses"
author: "Cesaire Tobias"
date: "2026-09-29"
format:
  pdf:
    date-format: "DD MMMM, YYYY"
    toc: true
    toc-depth: 2
    number-sections: true
    colorlinks: true
    documentclass: article
    geometry: margin=1in
    fontsize: 11pt
    linestretch: 1.5
    cite-method: citeproc
    bibliography: ../references.bib
    csl: ../harvard-cite-them-right.csl
    link-citations: true
    fig-pos: 'H'
    tbl-pos: 'H'
    include-in-header:
      text: |
        \usepackage{float}
        \usepackage{booktabs}
        \floatplacement{table}{H}
---

```{r}
#| label: setup
#| include: false
#| message: false
#| warning: false

# Every number in this document comes from the files the three analysis
# scripts write, or from the committed counts, so it cannot drift from the
# analysis. Run `Rscript R/analyse.R`, `Rscript R/analyse-generated.R` and
# `Rscript R/analyse-training.R` from the repository root before rendering.
suppressPackageStartupMessages({
  library(dplyr)
  library(knitr)
})

root <- ".."
derived <- c("generated-vs-varieties", "generated-among-conditions",
             "rates-generated", "rates-training", "composite-training",
             "training-contrasts", "training-per-subset",
             "training-exploratory", "olmocr-topics", "composite-ratios")
paths <- file.path(root, "data", paste0(derived, ".csv"))
if (!all(file.exists(paths))) {
  stop("run R/analyse.R, R/analyse-generated.R and R/analyse-training.R ",
       "from the repository root first; missing: ",
       paste(derived[!file.exists(paths)], collapse = ", "))
}
d <- setNames(lapply(paths, readr::read_csv, show_col_types = FALSE), derived)

read_data <- function(name) {
  readr::read_csv(file.path(root, "data", name), show_col_types = FALSE)
}
frames <- read_data("frames.csv")
run <- read_data("generated-run.csv")
by_text <- read_data("counts-generated-by-text.csv")
crawl_files <- jsonlite::fromJSON(
  file.path(root, "data", "dolma", "common_crawl.json"))$per_file

CONDITIONS <- c("base-1.0", "base-0.7", "instruct-1.0", "instruct-0.7")
VARIETIES <- c("US", "GB", "IE", "AU", "ZA")
FAMILIES <- c("announce", "contrastive", "epigram", "reveal")

fmt_ratio <- function(x) sprintf("%.2f", x)
fmt_p <- function(p) ifelse(p < 0.001, "< 0.001", sprintf("%.3f", p))
ratio_ci <- function(ratio, lower, upper) {
  sprintf("%s [%s, %s]", fmt_ratio(ratio), fmt_ratio(lower), fmt_ratio(upper))
}
fmt_rate <- function(x) sprintf("%.1f", x)
pct <- function(x) sprintf("%.0f%%", 100 * x)
number_word <- function(n) {
  c("no", "one", "two", "three", "four", "five", "six", "seven", "eight",
    "nine", "ten")[n + 1]
}
big <- function(x) format(x, big.mark = ",")
poisson_bounds <- function(hits, words) {
  ci <- sapply(hits, function(h) stats::poisson.test(h)$conf.int)
  list(lower = ci[1, ] / words * 1e6, upper = ci[2, ] / words * 1e6)
}

# The fourteen frames Study 1 could count, which every comparison against it
# uses.
shared <- d$`rates-generated` %>% filter(source == "US") %>% pull(frame_id)

# Study 1, for the rationale and the comparison of random-effect SDs.
s1 <- d$`composite-ratios`

# Study 2.
vs <- d$`generated-vs-varieties`
among <- d$`generated-among-conditions`
vs_row <- function(v) vs[vs$minus == v, ]
supported <- vs$minus[vs$supports_h1]

# The run.
finish <- by_text %>%
  group_by(source) %>%
  summarise(truncated = sum(finish_reason == "length"),
            share_truncated = mean(finish_reason == "length"),
            .groups = "drop")
run_tbl <- run %>%
  inner_join(finish, by = "source") %>%
  mutate(source = factor(source, levels = CONDITIONS)) %>%
  arrange(source)

# Summed rates over the shared frames, with exact intervals.
composite14 <- d$`rates-generated` %>%
  filter(frame_id %in% shared) %>%
  group_by(source) %>%
  summarise(hits = sum(hits), words = first(words), .groups = "drop") %>%
  mutate(rate = hits / words * 1e6)
bounds <- poisson_bounds(composite14$hits, composite14$words)
composite14 <- composite14 %>%
  mutate(lower = bounds$lower, upper = bounds$upper,
         source = factor(source, levels = c(VARIETIES, CONDITIONS))) %>%
  arrange(source)
summed <- function(s) composite14$rate[composite14$source == s]

# Study 3.
tc <- d$`training-contrasts`
primary <- tc %>% filter(comparison == "primary")
without_partial <- tc %>% filter(comparison == "primary, without the partial frames")
secondary <- tc %>% filter(comparison == "secondary")
whole_mix <- tc %>% filter(comparison == "secondary, approximate")
ct <- d$`composite-training`
base_rate15 <- ct$rate[ct$source == "base-1.0"]
mix_rate <- base_rate15 / whole_mix$ratio
explore <- d$`training-exploratory`
olmocr <- d$`olmocr-topics`

# Family rates relative to US, over the shared frames.
family_rates <- bind_rows(
  d$`rates-generated`,
  d$`rates-training` %>% filter(source == "common_crawl")) %>%
  filter(frame_id %in% shared) %>%
  group_by(source, family) %>%
  summarise(rate = sum(hits) / first(words) * 1e6, .groups = "drop")
family_vs_us <- family_rates %>%
  inner_join(family_rates %>% filter(source == "US") %>%
               select(family, us = rate), by = "family") %>%
  mutate(ratio = rate / us)
fam <- function(s, f) {
  family_vs_us$ratio[family_vs_us$source == s & family_vs_us$family == f]
}

# Per frame: how American Study 1 found it, how far the base model exceeds
# its input on it, and what post-training does to it.
per_frame <- bind_rows(
  d$`rates-generated`,
  d$`rates-training` %>% filter(source == "common_crawl")) %>%
  filter(frame_id %in% shared) %>%
  select(source, frame_id, rate) %>%
  tidyr::pivot_wider(names_from = source, values_from = rate) %>%
  mutate(american = US / ((GB + IE + AU + ZA) / 4),
         over = `base-1.0` / common_crawl,
         post = `instruct-1.0` / `base-1.0`) %>%
  inner_join(select(frames, frame_id, query, family), by = "frame_id") %>%
  arrange(desc(over))
rho <- cor(per_frame$american, per_frame$over, method = "spearman")
top_over <- head(per_frame$frame_id, 3)
top_american <- per_frame %>% arrange(desc(american)) %>% head(5) %>% pull(frame_id)
least_american <- per_frame$frame_id[which.min(per_frame$american)]
never <- per_frame$frame_id[per_frame$`base-1.0` == 0]
produced <- per_frame %>% filter(!is.nan(post))
most_raised <- produced$frame_id[which.max(produced$post)]

# Frames with no hits in any condition, which leave the model among them.
no_hits <- d$`rates-generated` %>%
  filter(source %in% CONDITIONS) %>%
  group_by(frame_id) %>%
  summarise(hits = sum(hits), .groups = "drop") %>%
  filter(hits == 0) %>%
  pull(frame_id)
instruct_multiple <- 1 / among$ratio[grepl("instruct", among$minus)]
base_07 <- among[among$minus == "base-0.7", ]

# Contrastive frames where a variety beat US in Study 1.
contrastive_beaten <- per_frame %>%
  filter(family == "contrastive", GB > US | AU > US) %>%
  nrow()

# How widely the commonest contrastive frame spreads across texts.
spread <- by_text %>%
  group_by(source) %>%
  summarise(texts = sum(F09 > 0), most = max(F09), .groups = "drop")
spread_of <- function(s, col) spread[[col]][spread$source == s]

# Frame rates by whether the text reached the token limit.
family_of <- setNames(frames$family, frames$frame_id)
by_family <- by_text
for (f in FAMILIES) {
  by_family[[f]] <- rowSums(by_text[shared[family_of[shared] == f]])
}
length_tbl <- by_family %>%
  group_by(source, finish_reason) %>%
  summarise(texts = n(), across(all_of(FAMILIES), sum),
            words = sum(words), .groups = "drop") %>%
  mutate(across(all_of(FAMILIES), ~ .x / words * 1e6),
         per_text = words / texts,
         source = factor(source, levels = CONDITIONS)) %>%
  arrange(source, desc(finish_reason))
by_finish <- function(s, reason, f) {
  length_tbl[[f]][length_tbl$source == s & length_tbl$finish_reason == reason]
}

# Variation between the drawn common_crawl files.
crawl_hits <- rowSums(as.matrix(crawl_files$hits))
crawl_rate <- crawl_hits / crawl_files$words * 1e6
crawl_total <- sum(crawl_hits) / sum(crawl_files$words) * 1e6
crawl_loo <- (sum(crawl_hits) - crawl_hits) /
  (sum(crawl_files$words) - crawl_files$words) * 1e6
crawl_range <- quantile(crawl_rate, c(0.05, 0.95))

# The prose below states these in words; stop rather than let it drift.
variety_family <- function(f) {
  family_vs_us$ratio[family_vs_us$source %in% VARIETIES[-1] &
                       family_vs_us$family == f]
}
stopifnot(
  length(intersect(top_over, top_american)) == 0,
  all(family_of[top_over] == "contrastive"),
  abs(crawl_total - ct$rate[ct$source == "common_crawl"]) < 1e-6,
  all(run$passes == 1),
  vs_row("US")$ratio_lower < 1, vs_row("US")$ratio_upper > 2,
  all(instruct_multiple > 1.8 & instruct_multiple < 2.2),
  fam("base-1.0", "epigram") < min(variety_family("epigram")),
  fam("base-1.0", "reveal") < max(variety_family("reveal")),
  fam("base-1.0", "reveal") > min(variety_family("reveal")),
  all(family_vs_us$ratio[family_vs_us$source == "common_crawl"] < 1)
)
```

# What this measures

Study 1 found that American web English in 2012 used fifteen frozen rhetorical
frames more often than the other four national varieties measured, which used
them at between `r fmt_ratio(min(s1$ratio))` and `r fmt_ratio(max(s1$ratio))`
times the American rate. The frames were chosen because they seem typical of
machine-generated prose: announcing a point before making it, the *it's not X,
it's Y* contrast, the epigrammatic assertion. The argument behind the whole
study is that such prose reads as foreign to readers of other varieties partly
because it reads as American.

Two further studies test the next steps. Each was registered before its own
data existed.

- **Study 2** (<https://osf.io/qjgtc>) asks whether a language model's
  unprompted prose sits at the American end of Study 1's spread. H1 predicts
  that the model's composite rate over the fourteen frames Study 1 could count
  is higher than each of GB, IE, AU and ZA. The comparison against US is
  reported without a predicted direction.
- **Study 3** (<https://osf.io/ngt3m>) asks whether the model uses the frames
  more often than the text it was trained on. A model that reproduces its
  training data's register is working correctly, so H1 predicts amplification:
  a composite rate above that of the web portion of the pretraining corpus.

The model is Olmo 3 7B, chosen because Ai2 published the corpus it was trained
on, which is what makes Study 3 possible. Four conditions were generated: the
base and instruction-tuned models
([`allenai/Olmo-3-1025-7B`](https://huggingface.co/allenai/Olmo-3-1025-7B),
[`allenai/Olmo-3-7B-Instruct`](https://huggingface.co/allenai/Olmo-3-7B-Instruct)),
each at unmodified sampling (temperature 1.0, top-p 1.0) and at typical
deployment settings (0.7 and 0.9). The confirmatory condition in both studies is
the base model at unmodified sampling, `base-1.0`. Each condition wrote one post
for each of 4,000 titles drawn from GloWbE's blog pages [@davies2013glowbe], 800
from each variety, prompted with `Write a blog post titled:` and the title, and
nothing else.

Four deviations were declared before the data they affect existed, and each was
filed as a registration update; all four are in `docs/deviations.md`. Two concern
the generated text. The instruct model receives the default system prompt its
chat template inserts (S2-D1), and every condition covers all 4,000 titles even
where that takes it past the registered two million words (S2-D2). The other two
concern the Dolma draw (S3-D1 and S3-D2).

# The generated text

```{r}
#| label: tbl-run
#| echo: false
run_tbl %>%
  transmute(Condition = as.character(source),
            Generations = big(generations),
            Words = big(words),
            `Reached 1,024 tokens` = sprintf("%s (%s)", big(truncated),
                                             pct(share_truncated)),
            Empty = empty_generations) %>%
  kable(caption = "The four conditions as generated on 28 September 2026.")
```

Each condition passed two million words on its first pass over the titles, so
no title was reused and topic is the same across all four. Between
`r pct(min(run_tbl$share_truncated))` and `r pct(max(run_tbl$share_truncated))`
of generations reached the 1,024-token limit and end mid-text. They are counted
as they came, as registered, and the length analysis below shows what that
does to the rates. The two empty generations are posts the base model ended
before writing anything. The checks made on the text before any frame was
counted are in `docs/collection-log.md`.

# Study 2: the model against the national varieties

## The confirmatory test

```{r}
#| label: tbl-confirmatory
#| echo: false
vs %>%
  transmute(Variety = minus,
            `base-1.0 against the variety` = ratio_ci(ratio, ratio_lower,
                                                      ratio_upper),
            `Holm p` = fmt_p(p_holm),
            `Supports H1` = ifelse(minus == "US", "no direction predicted",
                                   ifelse(supports_h1, "yes", "no"))) %>%
  kable(caption = paste("Confirmatory result: the composite rate of the",
                        length(shared), "shared frames in base-1.0, relative",
                        "to each variety, with unadjusted 95% intervals."))
```

H1 is supported against `r paste(supported, collapse = ", ")` only, where the
model's rate is `r fmt_ratio(vs_row("ZA")$ratio)` times South Africa's with a
Holm-adjusted p of `r fmt_p(vs_row("ZA")$p_holm)`. Against GB, IE and AU the
point estimates are above 1 and the adjusted p-values are not below 0.05. The
intervals in the table are unadjusted, which is why Ireland's excludes 1 while
its adjusted p is `r fmt_p(vs_row("IE")$p_holm)`. Against American English the
ratio is `r fmt_ratio(vs_row("US")$ratio)`, with an interval that runs from
well below the American rate to more than twice it.

The intervals are wide because the frames disagree about the model. The
frame-by-source SD is `r sprintf("%.2f", vs$sd_frame_source[1])` on the log
scale, against `r sprintf("%.2f", s1$sd_frame_variety[1])` among the varieties
in Study 1, so the model departs from the varieties frame by frame far more
than the varieties depart from each other. The exploratory section below shows
where.

```{r}
#| label: tbl-summed
#| echo: false
composite14 %>%
  transmute(Source = as.character(source),
            Hits = big(hits),
            `Per million` = fmt_rate(rate),
            `95% interval` = sprintf("[%s, %s]", fmt_rate(lower),
                                     fmt_rate(upper))) %>%
  kable(caption = paste("The", length(shared), "shared frames summed, for",
                        "each variety and each condition."))
```

Summed, base-1.0's rate is `r fmt_ratio(summed("base-1.0") / summed("US"))`
times the American one. The model's ratio of `r fmt_ratio(vs_row("US")$ratio)`
is smaller because it averages over frames on the log scale, where a large
excess on a few frames counts for less than it does in a sum.

## The four conditions

```{r}
#| label: tbl-among
#| echo: false
among %>%
  transmute(Comparison = sprintf("%s against %s", plus, minus),
            Ratio = ratio_ci(ratio, ratio_lower, ratio_upper),
            `Holm p` = fmt_p(p_holm)) %>%
  kable(caption = paste("The registered comparisons among the conditions,",
                        "Holm-corrected within the set of three."))
```

These comparisons use `r nrow(frames) - length(no_hits)` frames:
`r paste(no_hits, collapse = ", ")` have no hits in any condition and leave the
model, as registered. The instruction-tuned model uses the frames at about twice
the base model's rate, at either setting. This matches @reinhart2025llms, who
compared Llama 3 and GPT-4o with human writing on Biber's grammatical and
rhetorical features and found the differences larger for instruction-tuned
models than for base models. The base model at deployment settings
sits below the same model at unmodified sampling: base-1.0 is
`r ratio_ci(base_07$ratio, base_07$ratio_lower, base_07$ratio_upper)` times
base-0.7, with an adjusted p of `r fmt_p(base_07$p_holm)`. Both instruct
conditions include the default system prompt, so the difference between base
and instruct includes whatever that prompt contributes.

# Study 3: the model against its training data

## The primary test

The input is a 200-million-word sample of `common_crawl`, the web portion of
Dolma 3's stage-1 mix and 76% of its tokens by Ai2's published composition,
drawn as registered and recorded in `docs/collection-log.md`.

```{r}
#| label: tbl-primary
#| echo: false
bind_rows(primary, without_partial) %>%
  transmute(Frames = c("All fifteen", "Without F10 and F13"),
            `base-1.0 against common_crawl` = ratio_ci(ratio, ratio_lower,
                                                       ratio_upper),
            p = fmt_p(p_value)) %>%
  kable(caption = paste("Primary result, one uncorrected test, and its",
                        "registered robustness check."))
```

H1 is not supported. The confirmatory condition's rate is
`r fmt_ratio(primary$ratio)` times the input's, with an interval from
`r fmt_ratio(primary$ratio_lower)` to `r fmt_ratio(primary$ratio_upper)` and
p = `r fmt_p(primary$p_value)`. The data are consistent both with the model
reproducing its input's rate and with it exceeding that rate several times
over. Removing the two partial frames leaves the conclusion where it is. The
frame-by-source SD is `r sprintf("%.2f", primary$sd_frame_source)`, larger than
Study 2's, and again it is the frames' disagreement that makes the interval
wide.

## The other conditions

```{r}
#| label: tbl-secondary
#| echo: false
secondary %>%
  transmute(Condition = plus,
            `Against common_crawl` = ratio_ci(ratio, ratio_lower, ratio_upper),
            `Holm p` = fmt_p(p_holm)) %>%
  kable(caption = "Secondary results, Holm-corrected within the set of three.")
```

Both instruct conditions sit about three times above the input, with adjusted
p-values of `r fmt_p(max(secondary$p_holm[grepl("instruct", secondary$plus)]))`.
The base model at deployment settings is close to the input's rate.

## The whole mix

The registered comparison against the whole stage-1 mix gives
`r ratio_ci(whole_mix$ratio, whole_mix$ratio_lower, whole_mix$ratio_upper)`.
It rests on an approximation the registration names, token shares applied to
per-word rates, and it measures something different from the primary test. It
is a ratio of summed rates, base-1.0's `r fmt_rate(base_rate15)` per million
against a token-weighted mix rate of `r fmt_rate(mix_rate)`, and its interval
reflects counting noise alone. Frame disagreement, which widens every other
interval in this document, does not enter it, and a sum is dominated by the
frames with the most hits, which here are the contrastive frames the model
over-uses. The mix rate is also pulled down by the code, mathematics, arXiv,
science PDF and Wikipedia subsets, which use the frames far less than web text
does.

```{r}
#| label: tbl-subsets
#| echo: false
d$`training-per-subset` %>%
  inner_join(ct %>% select(minus = source, rate), by = "minus") %>%
  bind_rows(tibble(minus = "common_crawl",
                   rate = ct$rate[ct$source == "common_crawl"],
                   ratio = primary$ratio, ratio_lower = primary$ratio_lower,
                   ratio_upper = primary$ratio_upper)) %>%
  arrange(desc(rate)) %>%
  transmute(Subset = minus,
            `Per million` = fmt_rate(rate),
            `base-1.0 against the subset` = ratio_ci(ratio, ratio_lower,
                                                     ratio_upper)) %>%
  kable(caption = paste("Each subset's summed rate over all fifteen frames,",
                        "and the confirmatory condition against it, with",
                        "unadjusted intervals."))
```

The olmOCR draw reached `r sum(olmocr$files_drawn > 0)` of the subset's
`r nrow(olmocr)` topics. The ones it missed hold
`r pct(sum(olmocr$byte_share[olmocr$files_drawn == 0]))` of its compressed
bytes, health and education most of that. Reweighting the per-topic rates to
byte shares over the topics reached moves olmOCR's rate from
`r fmt_rate(explore$estimate[1])` to `r fmt_rate(explore$estimate[2])` per
million, and the whole-mix ratio from `r fmt_ratio(whole_mix$ratio)` to
`r fmt_ratio(explore$estimate[3])`. The olmOCR rate describes the unredacted
documents only (S3-D2).

# What drives the composite

Everything in this section is registered as exploratory. It is descriptive and
uncorrected, and no conclusion rests on a single frame.

## By family

```{r}
#| label: tbl-family
#| echo: false
family_vs_us %>%
  filter(source != "US") %>%
  mutate(source = factor(source, levels = c(VARIETIES, "common_crawl",
                                            CONDITIONS))) %>%
  select(source, family, ratio) %>%
  tidyr::pivot_wider(names_from = family, values_from = ratio) %>%
  arrange(source) %>%
  transmute(Source = as.character(source),
            across(all_of(FAMILIES), fmt_ratio)) %>%
  kable(caption = paste("Each family's summed rate over the shared frames,",
                        "relative to American English in 2012."))
```

The model differs from the varieties in shape as well as level. On the
announce-the-point frames, which separated the varieties most in Study 1,
base-1.0 is at `r fmt_ratio(fam("base-1.0", "announce"))` times the American
rate. On the epigrammatic frames it is at
`r fmt_ratio(fam("base-1.0", "epigram"))`, below every variety, and on the
single reveal frame at `r fmt_ratio(fam("base-1.0", "reveal"))`, among the
lower varieties. On the contrastive frames it is at
`r sprintf("%.1f", fam("base-1.0", "contrastive"))` times the American rate,
and the instruct model at `r sprintf("%.0f", fam("instruct-1.0", "contrastive"))`
times. The contrastive family separated the varieties least in Study 1, where
British or Australian writers used `r number_word(contrastive_beaten)` of its
four countable frames more often than Americans did.

The model's input, `common_crawl`, sits below American English of 2012 on every
family, near the other four varieties. Its rates rest on a different word count from Study 1's, and it
is a filtered crawl of English from everywhere, so the comparison is loose. It
does show that the stage-1 web text the model was trained on has no
contrastive excess of its own.

## Frame by frame

```{r}
#| label: tbl-frames
#| echo: false
per_frame %>%
  transmute(Frame = frame_id,
            Query = sprintf("`%s`", query),
            `US against the others, Study 1` = fmt_ratio(american),
            `base-1.0 against common_crawl` = fmt_ratio(over),
            `instruct-1.0 against base-1.0` = ifelse(is.nan(post), "no hits",
                                                     fmt_ratio(post))) %>%
  kable(caption = paste("The shared frames, ordered by how far base-1.0",
                        "exceeds its input. The first column is the US rate",
                        "over the mean of the other four varieties."))
```

The registration asks whether the frames the model most over-produces are the
ones Study 1 found most American. Across the `r nrow(per_frame)` frames, the
rank correlation between the first two ratio columns is
`r sprintf("%.2f", rho)`. The three frames the model over-produces most against
its input, `r paste(top_over, collapse = ", ")`, are all contrastive. None is
among Study 1's five most American frames, and `r least_american` is its least
American. The three frames the base model never produced,
`r paste(never, collapse = ", ")`, all ran at more than
`r sprintf("%.1f", floor(min(per_frame$american[per_frame$frame_id %in% never]) * 10) / 10)`
times the other varieties' rate in American English.

Post-training raises `r sum(produced$post > 1)` of the `r nrow(produced)`
frames the base model produced, `r most_raised` most of all.

## How widely the hits spread

A base model at unmodified sampling can fall into loops, and a few looping texts
would produce a large count. The contrastive hits are spread thinly instead. F09
appears in `r big(spread_of("base-1.0", "texts"))` of base-1.0's 4,000 texts and
never more than `r spread_of("base-1.0", "most")` times in one, and in
`r big(spread_of("instruct-1.0", "texts"))` of instruct-1.0's, at most
`r spread_of("instruct-1.0", "most")` in one.

## Length and truncation

```{r}
#| label: tbl-length
#| echo: false
length_tbl %>%
  transmute(Condition = as.character(source),
            Ending = ifelse(finish_reason == "length", "cut off", "finished"),
            Texts = big(texts),
            `Mean words` = sprintf("%.0f", per_text),
            across(all_of(FAMILIES), fmt_rate)) %>%
  rename_with(tools::toTitleCase, all_of(FAMILIES)) %>%
  kable(caption = paste("Family rates per million over the shared frames,",
                        "by whether the text finished or was cut off."))
```

Texts that reached the token limit were cut before their ending. If the frames
cluster at the ends of posts, truncation lowers a condition's rate. For the base
model it makes little difference: base-1.0's epigrammatic rate is
`r fmt_rate(by_finish("base-1.0", "stop", "epigram"))` per million in texts that
finished and `r fmt_rate(by_finish("base-1.0", "length", "epigram"))` in texts
that were cut, so its low epigrammatic rate is not an effect of truncation. The
instruct model's finished texts carry far more contrastive and epigrammatic
frames than its truncated ones. Truncation therefore lowers the instruct rates,
and the instruct conditions' excess is if anything understated.

## Variation between the drawn files

The 200 `common_crawl` files differ widely from one another: the middle 90% run
from `r fmt_rate(crawl_range[1])` to `r fmt_rate(crawl_range[2])` per million
over all fifteen frames. No single file matters much. Dropping any one of them
moves the subset's rate by at most
`r sprintf("%.1f%%", 100 * max(abs(crawl_loo / crawl_total - 1)))`.

# What this does not establish

- **One model.** Olmo 3 7B was chosen because its training corpus is published.
  Nothing here says how GPT, Claude or Gemini write, and for them the input side
  cannot be measured at all.
- **One prompt and one genre.** Every text answers `Write a blog post titled:`
  followed by a title from a 2012 blog. Other prompts and genres may draw on
  other registers.
- **Twelve years.** Study 1's baseline is web English collected in December
  2012, and the model card gives December 2024 as the cutoff of the model's
  data. A gap between the model and the varieties mixes register with whatever
  changed in web writing in between. Study 3's input comes from the model's own
  period.
- **Different denominators.** Study 1's rates divide by GloWbE's own word counts,
  and Studies 2 and 3 divide by this study's word rule (`docs/frame-mapping.md`).
  Comparisons within Studies 2 and 3 share one footing. Comparisons back to
  Study 1 do not, and the direction of any bias is unknown.
- **Fourteen frames against Study 1.** F06 is counted in Studies 2 and 3, but a
  free GloWbE account could not search it.
- **A system prompt in the instruct conditions.** The template inserts a default
  system prompt (S2-D1), so what post-training adds cannot be separated from
  what that prompt adds.
- **Stage 1 only.** The input sample comes from the stage-1 mix, 5.93 of the
  6.08 trillion tokens the
  [base model's card](https://huggingface.co/allenai/Olmo-3-1025-7B) lists
  across its three pretraining stages. The two later stages were not sampled,
  and they differ in kind: Ai2's
  [card for the midtraining mix](https://huggingface.co/datasets/allenai/dolma3_dolmino_mix-100B-1025)
  labels most of it synthetic, including instruction data and reasoning traces
  from other models, among them QwQ, Gemini and Llama Nemotron. The instruct
  model's post-training adds
  [supervised examples](https://huggingface.co/datasets/allenai/dolci-instruct-sft)
  that include responses written by GPT-4.1, and
  [preference pairs](https://huggingface.co/datasets/allenai/dolci-3-instruct-dpo-with-metadata)
  partly judged by a GPT model. The base model's departure from its stage-1
  input may therefore come partly from model-written text it saw later, and
  neither that departure nor the instruct model's excess can be traced to a
  stage.
- **An incomplete olmOCR draw.** It missed
  `r number_word(sum(olmocr$files_drawn == 0))` of the subset's topics and
  cannot see redacted documents. Both affect the whole-mix figure only.
- **Truncated texts.** Between `r pct(min(run_tbl$share_truncated))` and
  `r pct(max(run_tbl$share_truncated))` of generations were cut at 1,024 tokens.
  The length analysis above bounds what that does.
- **Fixed strings.** As in Study 1, these are rates per million. Density,
  sentence position and near-obligatory use are not measured.
- **Frames chosen as typical of model output.** A high model rate on some of
  them is what got them chosen. The studies ask where the model's rate sits
  against the varieties and against its input, which the choice does not settle,
  but a different fifteen frames could tell a different story.
- **Text that cannot be regenerated exactly.** Each generation's seed fixes its
  sampling request, but vLLM's arithmetic varies with the batch a request
  shares, so a rerun gives different text. `data/generated-sha256.txt`
  identifies the files that were counted.

# Reproducing this

The repository holds the generation code (`python/generate.py`), the titles
(`data/topics.csv`), the counts of the generated text
(`data/counts-generated.csv`, with each text's own counts in
`data/counts-generated-by-text.csv` and the run's totals in
`data/generated-run.csv`), the Dolma draw's per-file records (`data/dolma/`)
and their counts (`data/counts-dolma.csv` and `data/counts-dolma-topic.csv`),
and the record of both runs (`docs/collection-log.md`).

`Rscript R/analyse.R`, `Rscript R/analyse-generated.R` and
`Rscript R/analyse-training.R` recompute every number in this document from
those files. `python python/count_generated.py data/generated` recounts the
generated text into the tables above.

# References

::: {#refs}
:::
